---
title: "Parallelizing computation for score tests with radEmu"
author: "Sarah Teichman"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parallelizing computation for score tests with radEmu}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

First, we will install `radEmu`, if we haven't already. 

```{r, eval = FALSE}
# if (!require("remotes", quietly = TRUE))
#     install.packages("remotes")
# 
# remotes::install_github("statdivlab/radEmu")
```

Next, we can load `radEmu` as well as the `tidyverse` package suite. 

```{r setup, message = FALSE}
library(magrittr)
library(dplyr)
library(ggplot2)
library(stringr)
library(radEmu)
```

Finally, we will load the package `parallel`. 

```{r, message = FALSE}
library(parallel)
```

## Introduction 

In this vignette we will introduce parallel computing in order to do more efficient computation for score tests. Our recommendation for testing statistical hypotheses with small to moderate sample sizes with `radEmu` is to run a robust score test. While this test performs well (we like that it controls Type I error rate even in small samples!), it takes some time to run, because we need to fit our model under each null hypothesis. For differential abundance analysis, we often want to run a hypothesis test for each category (taxon, gene, etc) that we care about, so this adds up quickly. 
In order to improve computational efficiency, we can run these score tests in parallel using the `parallel` R package. This will let us take advantage of having additional cores on personal computers or computing clusters. Note that we will be using the `mclapply` function from the `parallel` package, which works on Mac and Linux machines, but does not on Windows. If you are using a Windows machine and would like a vignette about parallel computing on Windows, please let us know by opening an issue. 

We recommend that before working through this vignette, you start the an introduction to `radEmu` in "intro_radEmu.Rmd." 

## Setting up our radEmu model and running a single score test

We'll use the same Wirbel et al. data as in the introduction vignette. Recall that the [dataset published by Wirbel et al. (2019)](https://www.nature.com/articles/s41591-019-0406-6) is from a meta-analysis of case-controls comparing participants with and without colorectal cancer.

```{r}
# load in sample data
data("wirbel_sample")
# set group to be a factor with levels CTR for control and CRC for cancer
wirbel_sample$Group <- factor(wirbel_sample$Group, levels = c("CTR","CRC"))
# load in abundance data 
data("wirbel_otu")
# save mOTU names
mOTU_names <- colnames(wirbel_otu)
# consider taxa in the following genera
chosen_genera <- c("Eubacterium", "Faecalibacterium", "Fusobacterium", "Porphyromonas")
# get taxonomy information from mOTU names
mOTU_name_df <- data.frame(name = mOTU_names) %>% 
  mutate(base_name = stringr::str_remove(mOTU_names, "unknown ") %>%
                      stringr::str_remove("uncultured ")) %>%
  mutate(genus_name = stringr::word(base_name, 1))
# restrict to names in chosen genera
restricted_mOTU_names <- mOTU_name_df %>%
  filter(genus_name %in% chosen_genera) %>%
  pull(name)
# pull out observations from a chinese study within the meta-analysis
ch_study_obs <- which(wirbel_sample$Country %in% c("CHI"))
```

Now that we've processed our data, we can fit the `radEmu` model. Here we just want to get estimates for our parameters and their standard errors, but we will avoid running score tests by setting `run_score_tests = FALSE`.

```{r}
ch_fit <- emuFit(formula = ~ Group, 
                 data = wirbel_sample[ch_study_obs, ],
                 Y = wirbel_otu[ch_study_obs, restricted_mOTU_names],
                 run_score_tests = FALSE) 
```

In the introduction vignette we found that a meta-mOTU "unknown Eubacterium [meta_mOTU_v2_7116]" assigned to Eubacteria has a much higher ratio of abundance (comparing CRC group to control) than is typical across the mOTUs we included in this analysis, based on the parameter estimates in `ch_fit`. We can run a robust score test to test whether the differential abundance of this mOTU between cases and controls is significantly different from the differential abundance of a typical mOTU in our analysis. 

In order to run a single robust score test, we will set `run_score_tests = TRUE` and include the argument `test_kj`. Instead of re-estimating parameters in our model, we will provide `ch_fit` to the argument `fitted_model` and set `refit = FALSE`. 

```{r}
mOTU_to_test <- which(str_detect(restricted_mOTU_names, "7116"))
ch_fit$B %>% rownames
covariate_to_test <- which("GroupCRC" == ch_fit$B %>% rownames)
robust_score <- emuFit(formula = ~ Group,
                       data = wirbel_sample[ch_study_obs, ],
                       fitted_model = ch_fit,
                       refit = FALSE,
                       test_kj = data.frame(k = covariate_to_test, 
                                            j = mOTU_to_test), 
                       Y = as.matrix(wirbel_otu[ch_study_obs, restricted_mOTU_names]))
robust_score$coef$pval
```

Now, we can see that it took a little while to run our robust score test. If we investigate the coefficient table in our `robust_score` output, we can see that we have a p-value of `R robust_score$coef$pval[mOTU_to_test]` from our test. 

## Running robust score tests in parallel

Now, let's run some tests in parallel. We will be parallelizing our code over $j$ in the argument `test_kj`. We will assume that you have one covariate that you want to test, corresponding with a specific column of your design matrix $k$. However, if you want to run tests for multiple columns of your design matrix, then you can parallelize over pairs of $k$ and $j$ in the argument $test_kj$. 

Let's say that I want to run score tests for the first five mOTUs in our dataset. First, I need to check the cores on my computer to see a reasonable amount of cores to parallelize over. I tend to use one fewer core than the number of cores that I have available. 

```{r}
ncores <- parallel::detectCores() - 1
ncores
```

Next, I will write a function that will be called in parallel. This function will run my robust score test on the correct category. 

```{r}
emuTest <- function(category) {
  score_res <- emuFit(formula = ~ Group,
                       data = wirbel_sample[ch_study_obs, ],
                       fitted_model = ch_fit,
                       refit = FALSE,
                       test_kj = data.frame(k = covariate_to_test, 
                                            j = category), 
                       Y = as.matrix(wirbel_otu[ch_study_obs, restricted_mOTU_names]))
  return(score_res)
}
```

Now, we can run our score tests in parallel. This may take a minute or so. 

```{r}
score_res <- mclapply(1:5,
                      emuTest,
                      mc.cores = ncores)
```

Now, we can see that this barely took more time than running a single score test, because we were able to parallelize over cores (on my laptop, I'm using more than five cores, so I can run all five tests at the same time). The results are saved in a list. 

```{r}
score_res[[1]]$coef$pval
score_res[[2]]$coef$pval
```

Finally, we can make a coefficient matrix that combines the information from each component in our list. 

```{r}
full_score <- sapply(1:length(score_res), 
                     function(x) score_res[[x]]$coef$score_stat[x])
full_pval <- sapply(1:length(score_res), 
                     function(x) score_res[[x]]$coef$pval[x])
full_coef <- ch_fit$coef %>%
  dplyr::select(-score_stat, -pval) %>%
  filter(category_num %in% 1:5) %>%
  mutate(score_stat = full_score,
         pval = full_pval)
full_coef
```

