---
title: "Quick start with `radEmu`"
author: "Sarah Teichman and Amy Willis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup

First, we will install `radEmu`, if we haven't already. 

```{r, eval = FALSE}
# if (!require("remotes", quietly = TRUE))
#     install.packages("remotes")
# 
# remotes::install_github("statdivlab/radEmu")
```

Next, we can load `radEmu`.

```{r setup, message = FALSE}
library(radEmu)
```

In this vignette we'll explore a [dataset published by Wirbel et al. (2019)](https://www.nature.com/articles/s41591-019-0406-6). This is a meta-analysis of case-control studies of colorectal cancer, meaning that Wirbel et al. collected raw sequencing data from studies other researchers conducted and re-analyzed it.

Wirbel et al. published two pieces of data we'll focus on:

* metadata giving demographics and other information about participants
* a mOTU (metagenomic OTU) table

```{r}
data("wirbel_sample")
dim(wirbel_sample)
data("wirbel_otu")
dim(wirbel_otu)
```

We can see that we have $566$ samples and $845$ mOTUs. We will subset to samples from the Chinese study.

```{r}
wirbel_sample_ch <- wirbel_sample[wirbel_sample$Country == "CHI", ]
wirbel_otu_ch <- wirbel_otu[rownames(wirbel_sample_ch), ]
to_rm <- which(colSums(wirbel_otu_ch) == 0)
wirbel_otu_ch <- wirbel_otu_ch[, -to_rm]
```

## Differential abundance analysis with radEmu

```{r}
tmp <- emuFit(Y = wirbel_otu_ch, formula = ~ Group + Gender, data = wirbel_sample_ch,
              run_score_tests = FALSE, tolerance = 1e-3)
tmp1 <- emuFit(Y = wirbel_otu_ch, formula = ~ Group + Gender, data = wirbel_sample_ch,
              refit = FALSE, fitted_model = tmp, compute_cis = FALSE, verbose = "development", test_kj = data.frame(k = 2, j = 3))
```

