---
title: "Using radEmu on clustered data"
author: "Sarah Teichman"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using radEmu on clustered data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

First, we will install `radEmu`, if we haven't already. 

```{r, eval = FALSE}
# if (!require("remotes", quietly = TRUE))
#     install.packages("remotes")
# 
# remotes::install_github("statdivlab/radEmu")
```

Next, we can load `radEmu` as well as the `tidyverse` package suite. 

```{r setup, message = FALSE}
library(magrittr)
library(dplyr)
library(ggplot2)
library(stringr)
library(radEmu)
```

## Introduction 

In this vignette we will explore the use of `radEmu` with clustered data. Data with cluster dependence is a common phenomenon in microbiome studies. This type of dependence can come from experimental factors such as shared cages or tanks for study animals. When cluster dependence is not accounted for, statistical methods often operate under the assumption that all samples are independent. This will typically lead to anti-conservative inference (i.e. p-values that are smaller than they should be). 

Luckily, we have tools to account for cluster dependence in statistical inference, and they are implemented in `radEmu` through the `cluster` argument! This argument is only implemented from `radEmu` v1.2.0.0 forward, so if you are having trouble using the `cluster` argument, check that you have a recent enough version so use this functionality. 

## Generating data with cluster dependence

To start, let's generate a small data example in which there is cluster dependence within our data. 

```{r}
set.seed(10)
# 6 categories
J <- 6
# 2 columns in the design matrix 
p <- 2
# 60 samples 
n <- 60
# generate design matrix 
X <- cbind(1, rnorm(n))
cov_dat <- data.frame(cov = X[, 2])
# sample-specific effects
z <- rnorm(n) + 5
# cluster membership 
cluster <- rep(1:4, each = 15)
cov_dat$cluster <- cluster
# cluster effects 
cluster_effs <- lapply(1:4, function(i) log(matrix(rexp(2*J), nrow = 2)))
# intercepts for each category
b0 <- rnorm(J)
# coefficients for X1 for each category 
b1 <- seq(1, 5, length.out = J)
# mean center the coefficients
b1 <- b1 - mean(b1)
# set the coefficient for the 3rd category to 0
b1[3] <- 0
# generate B coefficient matrix 
b <- rbind(b0, b1)

# set up response matrix 
Y <- matrix(0, ncol = J, nrow = n)
for (i in 1:n) {
  for(j in 1:J){
    # mean model is exp(X_i %*% B_j + cluster_effect + z_i)
    temp_mean <- exp(X[i, , drop = FALSE] %*% 
                       (b[, j, drop = FALSE] + 
                          cluster_effs[[ cluster[i] ]][,j]) + z[i])
    # draw from a zero-inflated negative binomial with our mean 
    Y[i,j] <- rnbinom(1, mu = temp_mean, size = 5) * rbinom(1, 1, 0.8)
  }
}
```

Now that we have our data, we can get estimates of our parameters using the radEmu model and do inference. We will specifically test the $\beta_1$ parameter for the 3rd category, because we know that the true $\beta_1^3 = 0$. 

```{r}
ef_no_cluster <- emuFit(formula = ~ cov,
                        data = cov_dat, 
                        Y = Y,
                        test_kj = data.frame(k = 2, j = 3))
```

When we ignore clustering, we get an estimate of $\hat{\beta}_1^3$ of `r round(ef_no_cluster$coef$estimate[3], 3)` and a p-value of `r round(ef_no_cluster$coef$pval[3], 3)`.

```{r}
ef_cluster <- emuFit(formula = ~ cov,
                     data = cov_dat, 
                     Y = Y,
                     cluster = cluster, 
                     test_kj = data.frame(k = 2, j = 3))
```

Here, when we account for clustering, we get an estimate of  $\hat{\beta}_1^3$ of `r round(ef_cluster$coef$estimate[3], 3)` and a p-value of `r round(ef_cluster$coef$pval[3], 3)`. We can see that our estimates are the same whether or not we account for cluster, but our p-value is different.